---
title: "Excercise 2"
subtitle: "TMA4300 Computer intensive statistical methods Spring 2021"
author: 
- "Johan Fredrik Agerup"
- "Arne Rustad"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
header-includes:
   - \usepackage{subfig}
---

\newcommand{\vect}[1]{\ensuremath{\boldsymbol{\mathbf{#1} }}}
\newcommand{\matr}[1]{\ensuremath{\boldsymbol{\mathbf{#1} }}}
\newcommand{\Var}{\textrm{Var}}
\newcommand{\E}{\textrm{E}}
\newcommand{\Cov}{\textrm{Cov}}
\newcommand{\Corr}{\textrm{Corr}}

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3)
```


```{r rpackages,eval=TRUE,echo=FALSE}
library(ggplot2)
# library(expm)
# library(Matrix)
library(dplyr)
library(tidyverse)
library(boot)
library(stats)
library(latex2exp)
library(coda)
library(INLA)
library(mvtnorm)
library(MASS)
library(ggridges)
library(grDevices)
library(matrixcalc)
```

# Problem A



# Problem C


Let $x_1, \dots, x_n$ and $y_1, \dots, y_n$ be independent random variables where the $x_i$'s have an exponential distribution with $\lambda_0$ and the $y_i$'s have an exponential distribution with intensity $\lambda_1$. Assume we do not observe $x_1, \dots, x_n$ and $y_1, \dots, y_n$ directly, but that we observe

\begin{align*}
z_i &= \max \{x_i, y_i\} \qquad \textrm{for } i = 1, \dots, n \\
u_i &= I(x_i \geq y_i) \qquad \textrm{for } i = 1, \dots, n 
\end{align*}


where $I(A) = 1$ if $A$ is true and $0$ otherwise. $Thus for each $i = 1, \dots, n$ we observe the largest value of $x_i$ and $y_i$ and we know whether the observed value is $x_i$ or $y_i$. Based on the observed $(z_i, u_i) \qquad \textrm{for } i = 1, \dots, n $ we will in this problem use the EM algorithm to find the maximum likelihood estimates for $(\lambda_0, \lambda_1)$.


## a)

Define the following notation:

\begin{align*}
\vect{x} = [x_1, \dots, x_n] \\
\vect y = [y_1, \dots, y_n] \\
\vect z = [z_1, \dots, z_n] \\
\vect u = [u_1, \dots, u_n]
\end{align*}

The likelihood function for the complete data $(\vect x, \vect y)$ is

\begin{align*}
  L(\lambda_0, \lambda_1 | \vect x, \vect y) 
  &= f(\vect x, \vect y | \lambda_0, \lambda_1) \\
  &= f(\vect x | \lambda_0) f(\vect y | \lambda_1) \\
  &= \left [ \prod_{i=1}^n f(x_i | \lambda_0) \right ] \left [ \prod_{i=1}^n f(y_i | \lambda_1) \right ]
  &= \left [ \prod_{i=1}^n \lambda_0 e^{- \lambda_0 x_i} \right ]
  \left [ \prod_{i=1}^n \lambda_1 e^{- \lambda_1 y_i} \right ] \\
  &= \lambda_0^n e^{-\lambda_0 \sum_{i=1}^n x_i} \lambda_1^n e^{-\lambda_1 \sum_{i=1}^n y_i}.
\end{align*}

Correspondingly, the log likelihood function for the complete data $(\vect x, \vect y)$ is

\begin{equation*}
\begin{split}
l(\lambda_0, \lambda_1 | \vect x, \vect y) &= \ln \left( L(\lambda_0, \lambda_1 | \vect x, \vect y)  \right) \\
&= n \left( \ln(\lambda_0) + \ln(\lambda_1) \right) -\lambda_0 \sum_{i=1}^n x_i -\lambda_1 \sum_{i=1}^n y_i
\end{split}
\end{equation*}

Using this we want to find the expectation of the log likelihood function for the complete data condtional on the observed data $(\vect z, \vect u)$ and the estimates for the parameters at iteration $t$, i.e $(\lambda_o^{(t)}, \lambda_1^{(t)})$.

\begin{equation} \label{eq:em_formula_incomplete}
\begin{split}
\E \left[ l(\lambda_0, \lambda_1 | \vect x, \vect y)  | \vect z, \vect u, \lambda_0^{(t)}, \lambda_1^{(t)}\right]
&= \E \left[ \ln\left( f(\vect x, \vect y | \lambda_0, \lambda_1) \right) 
| \vect z, \vect u, \lambda_0^{(t)}, \lambda_1^{(t)}  \right] \\
&= n \left( \ln(\lambda_0) + \ln(\lambda_1) \right) -\lambda_0 \sum_{i=1}^n \E [x_i | z_i, u_i, \lambda_0^{(t)}]  - \lambda_1 \sum_{i=1}^n \E [y_i | z_i, u_i, \lambda_1^{(t)}
\end{split}
\end{equation}

For the case $u_i = 1$ then we have

\begin{align*}
\E [x_i | z_i, u_i = 1, \lambda_0^{(t)}] &= u_i
\end{align*}

and

\begin{align*}
\E [y_i | z_i, u_i = 1, \lambda_1^{(t)}
&= \int_0^{z_i} y_i P \left( y_i | y_i \leq z_i, \lambda_1^{(t)}\right) dy_i \\
&=\int_0^{z_i} y_i \frac{P \left( y_i \cap y_i \leq z_i | \lambda_1^{(t)} \right) }
{P \left( y_i \leq z_i | \lambda_1^{(t)} \right) } dy_i \\
&= \int_0^{z_i} y_i \frac{f(y_i | \lambda_1^{(t)})}{ \int_0^{z_i} f(y_i^* | \lambda_1^{(t)}) dy_i^*} dy_i) \\
&= \frac{1}{ \int_0^{z_i} \lambda_1^{(t)} e^{-\lambda_1^{(t)} y_i^* } dy_i^*}
\int_0^{z_i} y_i \lambda_1^{(t)} e^{-\lambda_1^{(t)} y_i} dy_i) \\
&= \frac{1}{1 - e^{- \lambda_1^{(t)} z_i}} 
\left( 
-z_i e^{- \lambda_1^{(t)} z_i} - \frac{1}{\lambda_1^{(t)}} e^{- \lambda_1^{(t)} z_i} + \frac{1}{\lambda_1^{(t)}}
\right) \\
&= \frac{-z_i e^{- \lambda_1^{(t)} z_i}}{1 - e^{- \lambda_1^{(t)} z_i}} + \frac{1}{\lambda_1^{(t)}} \\
&= \frac{1}{\lambda_1^{(t)}} - \frac{z_i}{e^{\lambda_1^{(t)} z_i} - 1}
\end{align*}

Correspondlingly, we have for the $u_i = 0$ that

\begin{align*}
\E [x_i | z_i, u_i = 0, \lambda_0^{(t)}
&= \int_0^{z_i} x_i P \left( x_i | x_i \leq z_i, \lambda_0^{(t)}\right) dx_i \\
&=\int_0^{z_i} x_i \frac{P \left( x_i \cap x_i \leq z_i | \lambda_0^{(t)} \right) }
{P \left( x_i \leq z_i | \lambda_0^{(t)} \right) } dx_i \\
&= \int_0^{z_i} x_i \frac{f(x_i | \lambda_0^{(t)})}{ \int_0^{z_i} f(x_i^* | \lambda_0^{(t)}) dx_i^*} dx_i) \\
&= \frac{1}{ \int_0^{z_i} \lambda_0^{(t)} e^{-\lambda_0^{(t)} x_i^* } dx_i^*}
\int_0^{z_i} x_i \lambda_0^{(t)} e^{-\lambda_0^{(t)} x_i} dx_i) \\
&= \frac{1}{1 - e^{- \lambda_0^{(t)} z_i}} 
\left( 
-z_i e^{- \lambda_0^{(t)} z_i} - \frac{1}{\lambda_0^{(t)}} e^{- \lambda_0^{(t)} z_i} + \frac{1}{\lambda_0^{(t)}}
\right) \\
&= \frac{-z_i e^{- \lambda_0^{(t)} z_i}}{1 - e^{- \lambda_0^{(t)} z_i}} + \frac{1}{\lambda_0^{(t)}} \\
&= \frac{1}{\lambda_0^{(t)}} - \frac{z_i}{e^{\lambda_0^{(t)} z_i} - 1}
\end{align*}

and

\begin{align*}
\E [y_i | z_i, u_i = 0, \lambda_1^{(t)}] &= u_i \\
\end{align*}

Inserting this result into Equation (\ref{eq:em_formula_incomplete}) we get

\begin{equation} \label{eq:em_formula}
\begin{split}
\E \left[ l(\lambda_0, \lambda_1 | \vect x, \vect y)  \big | \vect z, \vect u, \lambda_0^{(t)}, \lambda_1^{(t)}\right]
=& \E \left[ \ln\left( f(\vect x, \vect y | \lambda_0, \lambda_1) \right) 
\big | \vect z, \vect u, \lambda_0^{(t)}, \lambda_1^{(t)}  \right] \\
=& n \left( \ln(\lambda_0) + \ln(\lambda_1) \right) \\
&- \lambda_0 \sum_{i=1}^n 
\left [ u_i z_i 
+ (1 - u_i) \left( \frac{1}{\lambda_0^{(t)}} - \frac{z_i}{e^{\lambda_0^{(t)} z_i} - 1} \right)
\right] \\
&- \lambda_1 \sum_{i=1}^n \left [
(1 - u_i) z_i + 
u_i \left( \frac{1}{\lambda_1^{(t)}} - \frac{z_i}{e^{\lambda_1^{(t)} z_i} - 1} \right)
\right ]
\end{split}
\end{equation}

## 2)

Let $Q(\lambda_0, \lambda_1) be defined

\begin{equation*}

\end{equation*}
